---
title: 'Project: Public Health: National Nutritional Health'
author: "Goodness Nwokebu"
date: "2024-01-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

NHANES is a program run by the CDC to assess the health and nutritional status of adults and children in the US. It combines survey questions and physical examinations, including medical and physiological measurements and laboratory tests, and examines a representative sample of about 5,000 people each year. The data is used to determine the prevalence of diseases and risk factors, establish national standards, and support epidemiology studies and health sciences research. This information helps to develop public health policy, design health programs and services, and expand the nation's health knowledge.

Link to Dataset: <https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/nhanes.csv> 

### TASK 1: Process all NA (either by deleting or by converting to zero) 
```{r}
knitr::opts_chunk$set(echo = TRUE)
#Importing Necessary Libraries
library(tidyverse)


```


```{r}
#Importing Data and reading the data into a variable
url <- "https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/nhanes.csv"
Data <- read.table(url,header = TRUE,sep = ",")

cat("Number of rows:", nrow(Data), "\n")
cat("Number of columns:", ncol(Data), "\n")


#getting the summary statistics of the data
summary(Data)
```
```{r}
Data_tb = tibble(Data)
```

*Data Cleaning*

```{r fig.width=12}
#Diabetes and DiabetesAge Column
fd_tb <- Data_tb[
  complete.cases(Data_tb$Diabetes), #delete the nulls from diabetes column
  ] 
fd_tb %>% count(
  Diabetes, DiabetesAge) #check if the nullset in DiabetesAge aligns with No of Diabetes column
   
fd_tb$DiabetesAge[(fd_tb$Diabetes == "No")] <- 0 #equating those with no record of diabetes with zero diabetes years
fd_tb <- fd_tb[complete.cases(fd_tb$DiabetesAge),] #delete the remaining nulls from DiabetesAge column

cat("Null value in Diabetes Column:", sum(is.na(fd_tb$Diabetes)), "\n")
cat ("null value in DiabetesAge Column:", sum(is.na(fd_tb$DiabetesAge)), "\n")

#npregnanies Column
fd_tb$nPregnancies[is.na(fd_tb$nPregnancies)] <- 0 #equating null to having zero pregnancies.

#BMI, PULSE, BPSys, BPDia, Testerone
col_to_del = c("BMI", "Pulse", "BPSys", 'BPDia', "Testosterone", 'HDLChol','TotChol') #delete null col from BMI, Pulse, BPSy s, BPDia, Testosterone, HDLchol and TotChol
filteredata_tb <- subset(fd_tb, complete.cases(fd_tb[, col_to_del]))

#nBabies          
nBabies_NA <- subset(filteredata_tb, is.na(nBabies))
filteredata_tb %>% count(nPregnancies, nBabies) #getting the summary data of nPrgnancies and nBabies.

#filling null values in nBabies where pregnancy is 0 with 0.
filteredata_tb <- filteredata_tb %>% mutate( nBabies_filled = ifelse(nPregnancies ==0 & is.na(nBabies),0, nBabies)) 
filt_tb <- filteredata_tb %>% group_by( nPregnancies) %>% mutate(
  nBabies_filled = ifelse(is.na(nBabies_filled), round(mean(nBabies_filled, na.rm = TRUE), digits = 0), nBabies_filled)) #filling with the mean nBabies for each group of Pregnancies
filteredata_tb <- filteredata_tb %>% mutate(
  nBabies_filled = filt_tb$nBabies_filled) 

filteredata_tb$nBabies = NULL  #removing the nBabies column

#Alcoholday, AlcoholYear, Income, HomeRooms, SmokingStatus
filteredata_tb$AlcoholYear[is.na(filteredata_tb$AlcoholYear)] <- 0
filteredata_tb$AlcoholDay[is.na(filteredata_tb$AlcoholDay)] <- 0
filteredata_tb$Income[is.na(filteredata_tb$Income)] <- 0
filteredata_tb$HomeRooms[is.na(filteredata_tb$HomeRooms)] <-0
filteredata_tb$SmokingStatus[is.na(filteredata_tb$SmokingStatus)] <- "Never"

 

```
### TASK 2: Visualize the distribution of BMI, Weight, Weight in pounds (weight *2.2) and Age with an histogram.

```{r}
#Data Analysis
dataplot1 <- select(filteredata_tb,Weight, BMI, Age)
#computing weight in pounds
dataplot1 <- dataplot1 %>% mutate(Weightsp = Weight * 2.2)

#plotting thr histograms
plots <- list()

for (col_name in names(dataplot1)) {
  # Use .. notation to dynamically refer to the column in aes()
  plot <- ggplot(dataplot1, aes_string(x = col_name)) +
    geom_histogram(fill = "lightblue", col = 'black') +
    labs(title = paste("Histogram of", col_name))

  plots[[col_name]] <- plot
}

# Arrange and print the plots
gridExtra::grid.arrange(grobs = plots, ncol = 2) 


```
The four histogram shows the distribution of Weight, BMI, Age and the Weight(in pounds). Excerpts:
1) Most of the population's weight fell between 70 - 80kg and 140-180 pounds
2) Participants aged 60 and above were less represented in the population
3) A lot of the participants seems to be obese having BMI >30  

#### TASK 3: What’s the mean 60-second pulse rate for all participants in the data?

```{r}
round(mean(filteredata_tb$Pulse)) #mean pulse of the population


```
The population mean pulse is 74



#### TASK 4: What’s the range of values for diastolic blood pressure in all participants?


```{r}
paste("The range of diastolic blood pressure of all participant is", 
      min(filteredata_tb$BPDia), "-", max(filteredata_tb$BPDia))

```

#### TASK 5: What’s the variance and standard deviation for income among all participants?
```{r}
paste("The standard deviation and variance of the partcipants income  are", 
      round(sd(filteredata_tb$Income)), "and", round(var(filteredata_tb$Income)),
      "respectively")
```
This shows that there is a significant difference in the individual's income. 
This maybe from the fact that about 20% of the population are under eighteen years
with a no official means of income.

#### TASK 6: Visualize the relationship between weight and height 
#### Color the points by:
* gender
* diabetes
* smoking status
```{r}
#Colour by Gender
ggplot(filteredata_tb, aes(x = Height,y = Weight,
                           color = Gender))+ geom_point() + labs(
                             title = "Scatter Plot of Height and Weight",
                           x = "Height (cm)", y = "Weight (kg)")

#colour by diabetes
ggplot(filteredata_tb, aes(x = Height, y = Weight, 
                           color = Diabetes)) + geom_point() + labs(
                             title = "Scatter Plot of Height and Weight",
                          x = "Height (cm)", y = "Weight (kg)")

#Colour by Smoking Status
ggplot(filteredata_tb, aes(x = Height, y = Weight, 
                           color = SmokingStatus)) + geom_point() + labs(
                             title = "Scatter Plot of Height and Weight",                   
                             x = "Height (cm)", y = "Weight (kg)")

```

#### TASK 7: Conduct t-test between the following variables and make conclusions on the relationship between them based on P-Value
* Age and Gender
* BMI and Diabetes
* Alcohol Year and Relationship Status
```{r}
print(t.test(Age ~ Gender, data= filteredata_tb))
```

This test aims to check if the there is a significant statistical differences between the Female mean age and that of the males. From the result above, female group have an average mean of 41.72 while that of male is 41.11. A positive t-test was gotten meaning there is a positive which fully supports that the mean age of females is greater than male but it is not statistically significant(P-value > 0.05). This means that we do not have enough evidence to reject the null hypothesis. Hence the variability in the means can be random either than a true difference.


```{r}
print(t.test(BMI ~ Diabetes, data= filteredata_tb))

```
The mean BMI of people with Diabetes(Diabetes is equal to Yes) is higher than those without(Diabetes is equal to No). P_value of 2.2E-16 is lesser than 0.05 and the confidence interval not including 0 shows  that this difference is statistically significant and likely not by chance. 

```{r}
print(t.test(AlcoholYear~RelationshipStatus, data= filteredata_tb))
```
The small p-value and the confidence interval not including 0 suggest strong evidence that the mean alcohol consumption is different between individuals in the "Committed" and "Single" relationship statuses. The positive t-value indicates that, on average, individuals in the "Committed" group have a higher mean alcohol consumption than those in the "Single" group. The difference in means is estimated to be between 9.70 and 23.07 units.



